# Common Distributions

## Bernoulli Distribution

| Property         | Description                                                                 |
|------------------|-----------------------------------------------------------------------------|
| **Name**         | Bernoulli distribution                                                      |
| **Notation**     | $\text{Bernoulli}(\pi)$                                                     |
| **Parameters**   | $0 \le \pi \le 1$: probability of success                                   |
| **Support**      | $x \in \{0,1\}$                                                             |
| **PMF**          | $f(x) = \pi^x (1-\pi)^{1-x}$                                                |
| **CDF**          | $F(x) = 0$ for $x < 0$; $1-\pi$ for $0 \le x < 1$; $1$ for $x \ge 1$        |
| **Mean**         | $\pi$                                                                       |
| **Variance**     | $\pi(1-\pi)$                                                                |
| **R functions**  | `dbern(x, prob)` (PMF) <br> `pbern(q, prob)` (CDF) <br> `qbern(p, prob)` (quantile) <br> `rbern(n, prob)` (random sampling) <br> *(from **extraDistr** package)* |
| **Special cases**| $\text{Binomial}(n=1, \pi)$ is $\text{Bernoulli}(\pi)$. |

: Summary of the Bernoulli distribution {#tbl-bernoulli}


## Beta Distribution

| Property         | Description                                                                 |
|------------------|-----------------------------------------------------------------------------|
| **Name**         | Beta distribution                                                           |
| **Notation**     | $\text{Beta}(\alpha, \beta)$                                                |
| **Parameters**   | $\alpha > 0$: shape; $\beta > 0$: shape                                     |
| **Support**      | $x \in (0, 1)$                                                              |
| **PDF**          | $f(x) = \dfrac{x^{\alpha - 1} (1 - x)^{\beta - 1}}{B(\alpha, \beta)}$       |
| **CDF**          | $F(x) = I_x(\alpha, \beta)$, the regularized incomplete beta function       |
| **Mean**         | $\dfrac{\alpha}{\alpha + \beta}$                                            |
| **Variance**     | $\dfrac{\alpha \beta}{(\alpha + \beta)^2 (\alpha + \beta + 1)}$             |
| **R functions**  | `dbeta(x, alpha, beta)` (density) <br> `pbeta(q, alpha, beta)` (CDF) <br> `qbeta(p, alpha, beta)` (quantile) <br> `rbeta(n, alpha, beta)` (random sampling) <br> *(base R)* |
| **Special cases**| $\text{Beta}(1,1)$ is $\text{Uniform}(0,1)$. |

: Summary of the Beta distribution {#tbl-beta}


## Location-Scale *t* Distribution

| Property         | Description                                                                 |
|------------------|-----------------------------------------------------------------------------|
| **Name**         | Location-scale $t$ distribution                                             |
| **Notation**     | $t(\mu, \sigma, \nu)$                                                       |
| **Parameters**   | $\mu \in \mathbb{R}$: location; $\sigma > 0$: scale; $\nu > 0$: degrees of freedom |
| **Support**      | $x \in \mathbb{R}$                                                          |
| **PDF**          | $f(x) = \dfrac{\Gamma\left(\frac{\nu+1}{2}\right)}{\Gamma\left(\frac{\nu}{2}\right)\sqrt{\nu\pi}\sigma} \left[1 + \dfrac{1}{\nu} \left(\dfrac{x - \mu}{\sigma}\right)^2\right]^{-\frac{\nu+1}{2}}$ |
| **CDF**          | $F(x) = T_\nu\left(\dfrac{x - \mu}{\sigma}\right)$, where $T_\nu$ is the CDF of the standard $t$ distribution |
| **Mean**         | $\mu$ for $\nu > 1$; undefined for $\nu \le 1$                              |
| **Variance**     | $\dfrac{\nu \sigma^2}{\nu - 2}$ for $\nu > 2$; infinite for $1 < \nu \le 2$; undefined for $\nu \le 1$ |
| **R functions**  | `dt.scaled(x, df, mean = mu, sd = sigma)` (density) <br> `pt.scaled(q, df, mean = mu, sd = sigma)` (CDF) <br> `qt.scaled(p, df, mean = mu, sd = sigma)` (quantile) <br> `rt.scaled(n, df, mean = mu, sd = sigma)` (random sampling) <br> *(from **metRology** package)* |
| **Special cases**| - $t(0,1,\nu)$: standard Student's $t$  <br> - $\nu \to \infty$: converges to $\mathcal{N}(\mu, \sigma^2)$  <br> - Heavy-tailed alternative to the normal distribution |

: Summary of the location-scale $t$ distribution {#tbl-t}

```{r}
#| echo: false
#| message: false
#| warning: false
# Load necessary libraries
library(ggplot2)
library(metRology)
library(dplyr)
library(tidyr)

# Set parameters
mu <- 10
sigma <- 2
df_values <- c(1, 3, 5)

# Create a sequence of x values
x <- seq(0, 20, length.out = 500)

# Create a data frame for the t-distributions
t_dists <- expand.grid(x = x, df = df_values) |>
  mutate(density = dt.scaled(x, df = df, mean = mu, sd = sigma),
         distribution = paste0("t(df=", df, ")"))

# Create the normal distribution data
normal_df <- data.frame(
  x = x,
  density = dnorm(x, mean = mu, sd = sigma),
  distribution = "Normal"
)

# Combine both
plot_df <- bind_rows(t_dists, normal_df)

# Plot
ggplot(plot_df, aes(x = x, y = density, color = distribution)) +
  geom_line(linewidth = 1) +
  labs(
    title = "Location-scale t-distribution vs Normal distribution",
    x = "x",
    y = "Density",
    color = "Distribution"
  ) +
  theme_minimal()
```


## Negative Binomial Distribution

| Property         | Description                                                                 |
|------------------|-----------------------------------------------------------------------------|
| **Name**         | Negative binomial distribution                                              |
| **Notation**     | $\text{NB}(\mu, r)$                                                         |
| **Parameters**   | $\mu > 0$: mean <br> $r > 0$: size (dispersion)                             |
| **Support**      | $x \in \{0,1,2,\ldots\}$                                                    |
| **PMF**          | $f(x) = \dfrac{\Gamma(x+r)}{\Gamma(r)\,x!} \left(\dfrac{r}{r+\mu}\right)^r \left(\dfrac{\mu}{r+\mu}\right)^x$ |
| **CDF**          | $F(x) = \displaystyle \sum_{k=0}^{\lfloor x \rfloor} \dfrac{\Gamma(k+r)}{\Gamma(r)\,k!} \left(\dfrac{r}{r+\mu}\right)^r \left(\dfrac{\mu}{r+\mu}\right)^k$ |
| **Mean**         | $\mu$                                                                       |
| **Variance**     | $\mu + \dfrac{\mu^2}{r}$                                                    |
| **R functions**  | `dnbinom(x, size = r, mu = mu)` (PMF) <br> `pnbinom(q, size = r, mu = mu)` (CDF) <br> `qnbinom(p, size = r, mu = mu)` (quantile) <br> `rnbinom(n, size = r, mu = mu)` (random sampling) <br> *(base R)* |
| **Special cases**| As $r \to \infty$, $\text{NB}(\mu, r) \to \text{Poisson}(\mu)$. |

: Summary of the negative binomial distribution {#tbl-nbinom}

```{r}
#| echo: false
#| message: false
#| warning: false
library(ggplot2)

# Parameters
mu <- 3
r <- 2
x <- 0:15
df <- data.frame(
  x = x,
  prob = dnbinom(x, size = r, mu = mu)
)

ggplot(df, aes(x = x, y = prob)) +
  geom_col(fill = "steelblue") +
  geom_point(size = 2, color = "black") +
  labs(
    title = "Negative Binomial Distribution (μ = 3, r = 2)",
    x = "x",
    y = "Probability"
  ) +
  theme_minimal()
```


## Poisson Distribution

| Property         | Description                                                                 |
|------------------|-----------------------------------------------------------------------------|
| **Name**         | Poisson distribution                                                        |
| **Notation**     | $\text{Poisson}(\lambda)$                                                   |
| **Parameters**   | $\lambda > 0$: rate (mean number of events per unit interval)               |
| **Support**      | $x \in \{0,1,2,\ldots\}$                                                    |
| **PMF**          | $f(x) = \dfrac{e^{-\lambda} \lambda^x}{x!}$                                |
| **CDF**          | $F(x) = \displaystyle \sum_{k=0}^{\lfloor x \rfloor} \dfrac{e^{-\lambda}\lambda^k}{k!}$ |
| **Mean**         | $\lambda$                                                                   |
| **Variance**     | $\lambda$                                                                   |
| **R functions**  | `dpois(x, lambda)` (PMF) <br> `ppois(q, lambda)` (CDF) <br> `qpois(p, lambda)` (quantile) <br> `rpois(n, lambda)` (random sampling) <br> *(base R)* |
| **Special cases**| - Limit of $\text{Binomial}(n,p)$ as $n \to \infty$, $p \to 0$ with $np=\lambda$ fixed <br> - Distribution of counts in a homogeneous Poisson process <br> - Sum of independent $\text{Poisson}(\lambda_i)$ is $\text{Poisson}(\sum \lambda_i)$ |

: Summary of the Poisson distribution {#tbl-poisson}

```{r}
#| echo: false
#| message: false
#| warning: false
library(ggplot2)

# Set parameter
lambda <- 3
x <- 0:15
df <- data.frame(
  x = x,
  prob = dpois(x, lambda)
)

ggplot(df, aes(x = x, y = prob)) +
  geom_col(fill = "darkred") +
  geom_point(size = 2, color = "black") +
  labs(
    title = "Poisson Distribution (λ = 3)",
    x = "x",
    y = "Probability"
  ) +
  theme_minimal()
```
