# Ordered Logit

## Ordered Outcomes

Suppose an ordered outcome $y_i \in \{1, 2, \ldots, J\}$, where $\{1, 2, \ldots, J\}$ correspond to *qualitative labels* that have a natural ordering.

*Example 1.* The ANES asks respondents to place themselves along a 7-point **ideology** scale with positions labeled "extremely liberal," "liberal," "slightly liberal," "moderate/middle of the road," "slightly conservative," "conservative," and "extremely conservative." We typically let 1, 2, 3, 4, 5, 6, and 7. correspond to these seven ordered labels. 

*Example 2*. The **Political Terror Scale** (PTS) is a country-year measure of state repression. This measure assigns one of five qualitative labels summarizing the amount of state repression to each country-years based on reports from Amnesty International and the U.S. State Department.

| Numeric Coding      | Qualitative Label |
|------------|----------------|
| 1  | Countries under a secure rule of law, people are not imprisoned for their views, and torture is rare or exceptional. Political murders are extremely rare. |
| 2  | There is a limited amount of imprisonment for nonviolent political activity. However, few persons are affected, torture and beatings are exceptional. Political murder is rare. |
| 3  | There is extensive political imprisonment, or a recent history of such imprisonment. Execution or other political murders and brutality may be common. Unlimited detention, with or without a trial, for political views is accepted. |
| 4  | Civil and political rights violations have expanded to large numbers of the population. Murders, disappearances, and torture are a common part of life. In spite of its generality, on this level terror affects those who interest themselves in politics or ideas. |
| 5  | Terror has expanded to the whole population. The leaders of these societies place no limits on the means or thoroughness with which they pursue personal or ideological goals. |

## Ordered Logit

Let linear predictor $\eta_i$ be our usual linear predictor so that $\eta_i = X_i^\beta$, except *without an intercept $\beta_0$.*

Now define $J-1$ "cutpoints" $\alpha_1 < \alpha_2 < \cdots < \alpha_{J-1}$ that act as a separate intercept for each category.

The **ordered logit** defines the the *cumulative* probabilities as

$$
\Pr(Y_i \leq j) = \text{logit}^{-1}(\alpha_j - \eta_i),
\qquad j = 1, \ldots, J-1.
$$

Then probabilities for each category are

$$
\begin{align}
\Pr(Y_i = 1) &= \text{logit}^{-1}(\alpha_1 - \eta_i) \\
\Pr(Y_i = 2) &= \text{logit}^{-1}(\alpha_2 - \eta_i) - \text{logit}^{-1}(\alpha_{1} - \eta_i) \\
\Pr(Y_i = 3) &= \text{logit}^{-1}(\alpha_3 - \eta_i) - \text{logit}^{-1}(\alpha_{3} - \eta_i) \\
             &\vdots\\
\Pr(Y_i = j) &= \overbrace{\text{logit}^{-1}(\alpha_j - \eta_i)}^{\Pr(Y_i \leq j)} - \overbrace{\text{logit}^{-1}(\alpha_{j-1} - \eta_i)}^{^{\Pr(Y_i \leq j-1)}}, \qquad j = 2, \ldots, J-1 \\
             &\vdots\\
\Pr(Y_i = J) &= 1 - \text{logit}^{-1}(\alpha_{J-1} - \eta_i).
\end{align}
$$

The figure below shows this graphically for an outcome with four possible values.

- I set the three cutpoints to $\alpha = (-2.5,-1.3,1.3)$. 
- I set the linear predictor for an example observation to $\eta_i = -1.5$. 

The figure shows how $\alpha$ and $\eta_i$ are translated into $\Pr(y_i = j)$.

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-asp: 0.75
#| fig-width: 12
library(tidyverse)
library(latex2exp)


# parameters
alpha <- sort(c(-2.5, -1.3, 1.3))
J <- length(alpha) + 1
eta   <- -1.5   

# make a color pallette
library(RColorBrewer)
colors_J <- brewer.pal(J, "Dark2")  # or "Set1", "Paired", "Accent"

cumulative_prob <- c(plogis(alpha - eta), 1)
prob <- c(cumulative_prob[1], diff(cumulative_prob))

# S-shaped curve
curve_df <- tibble(x = seq(-8, 8, by = 0.001)) |>
  mutate(prob = plogis(x))

# lines
x_lines_df <- tibble(x0 = alpha - eta, x1 = alpha - eta, 
                     y0 = 0, y1 = cumulative_prob[1:(J - 1)])

y_lines_df <- tibble(x0 = min(curve_df$x), x1 = alpha - eta, 
                     y0 = cumulative_prob[1:(J - 1)], y1 = cumulative_prob[1:(J - 1)])

# labels
x_labs <- paste0(
  r"($\alpha_{)", 1:(J - 1), r"(} - \eta_i = )",
  scales::number(alpha - eta, accuracy = 0.01),
  r"($)"
)
x_labels_df <- tibble(
  x = alpha - eta,
  y = cumulative_prob[1:(J - 1)]/2,
  label = TeX(x_labs)
)
y_labs <- paste0(
  r"($\Pr(Y_i \leq )", 1:(J - 1), ")",
  r"(=logit^{-1}(\alpha_{)", 1:(J - 1),
  r"(} - \eta_i)=)",
  scales::number(cumulative_prob[1:(J - 1)], accuracy = 0.01),
  r"($)"
)
y_labels_df <- tibble(
  x = (min(curve_df$x) + (alpha - eta)) / 2,
  y = cumulative_prob[1:(J - 1)],
  label = TeX(y_labs), 
  region = as.character(1:(J-1))
)

# probability labels
probs_df <- tibble(x = min(curve_df$x), y = cumulative_prob)
prob_labs <- paste0(
  r"($\Pr(Y_i = )", 1:J, ") = ",
  r"(\Pr(Y_i \leq )", (1:J), ") - ",
  r"(\Pr(Y_i \leq )", 1:J - 1, ") = ", 
  scales::number(prob[1:J], accuracy = 0.01),
  r"($)"
)
prob_labs[1] <- paste0(
  r"($\Pr(Y_i = 1) = \Pr(Y_i \leq 1) = )", 
  scales::number(prob[1], accuracy = 0.01),
  r"($)"
)
prob_labs[J] <- paste0(
  r"($\Pr(Y_i = )", J, ") =", r"(1 - \Pr(Y_i \leq )", J - 1, ") =", 
  scales::number(prob[J], accuracy = 0.01),
  r"($)"
)
prob_labels_df <- tibble(
  x = min(curve_df$x),
  y = cumulative_prob - diff(c(0, cumulative_prob))/2,
  label = TeX(prob_labs), 
  region = as.character(1:(J))
)

# y polygons: use a for-loop to build the polygons one at a time
y_poly_list <- NULL
y_poly_list[[1]] <- curve_df |>
  filter(x <= alpha[1] - eta) |>
  mutate(to = cumulative_prob[1], 
         from = prob) |>
  mutate(region = as.character(1))
for (j in 2:(J - 1)) {
  y_poly_list[[j]] <- curve_df |>
    filter(x <= alpha[j] - eta) |>
    mutate(to = cumulative_prob[j], 
           from = ifelse(cumulative_prob[j-1] > prob, cumulative_prob[j - 1], prob)) |>
    mutate(region = as.character(j))
}
y_poly_list[[J]] <- curve_df |>
  mutate(to = 1, 
         from = ifelse(cumulative_prob[J-1] > prob, cumulative_prob[J - 1], prob)) |>
  mutate(region = as.character(J))
y_poly_df <- bind_rows(y_poly_list)

# x polygons: use a for-loop to build the polygons one at a time
x_poly_list <- NULL
x_poly_list[[1]] <- curve_df |>
  filter(x <= alpha[1] - eta) |>
  mutate(to = prob, 
         from = 0) |>
  mutate(region = as.character(1))
for (j in 2:(J - 1)) {
  x_poly_list[[j]] <- curve_df |>
    filter(x <= alpha[j] - eta, x > alpha[j - 1] - eta) |>
    mutate(to = prob, 
           from = 0) |>
    mutate(region = as.character(j))
}
x_poly_list[[J]] <- curve_df |>
  filter(x > alpha[J - 1] - eta) |>
  mutate(to = prob, 
         from = 0) |>
  mutate(region = as.character(J))
x_poly_df <- bind_rows(x_poly_list)

# probability blocks
width <- diff(range(curve_df$x))*0.05
block_list <- NULL
block_list[[1]] <- tibble(x = seq(min(curve_df$x) - width, min(curve_df$x), by = 0.01)) |>
  mutate(from = 0, to = cumulative_prob[1], region = "1") 
for (j in 2:(J - 1)) {
  block_list[[j]] <- tibble(x = seq(min(curve_df$x) - width, min(curve_df$x), by = 0.01)) |>
    mutate(from = cumulative_prob[j-1], to = cumulative_prob[j], region = as.character(j))
}
block_list[[J]] <- tibble(x = seq(min(curve_df$x) - width, min(curve_df$x), by = 0.01)) |>
  mutate(from = cumulative_prob[J - 1], to = cumulative_prob[J], region = as.character(J))


block_df <- bind_rows(block_list)

ggplot() +
  geom_ribbon(data = block_df, aes(x = x, ymin = from, ymax = to, fill = region), alpha = 1.0, color = "black") +
  geom_segment(aes(x = min(curve_df$x), y = 0, yend = 1), color = "black") + 
  geom_segment(aes(x = min(curve_df$x) - width, y = 0, yend = 1), color = "black") + 
  geom_ribbon(data = x_poly_df, aes(x = x, ymin = from, ymax = to, fill = region), alpha = 0.35) +
  geom_ribbon(data = y_poly_df, aes(x = x, ymin = from, ymax = to, fill = region), alpha = 0.35) +
  
  geom_segment(data = x_lines_df, aes(x = x0, xend = x1, y = y0, yend = y1), linetype = "dotted") +
  geom_label(data = x_labels_df, aes(x = x, y = y, label = label), parse = TRUE) +
  geom_segment(data = y_lines_df, aes(x = x0, xend = x1, y = y0, yend = y1), linetype = "dotted") +
  geom_segment(data = y_labels_df, aes(x = x, xend = x, y = y, yend = 0, color = region), linewidth = 4) + 
  
  geom_label(data = y_labels_df, aes(x = x, y = y, label = label, color = region), parse = TRUE) +
  geom_line(data = curve_df, aes(x = x, y = prob)) +
  geom_point(data = x_lines_df, aes(x = x1, y = y1), shape = 21, fill = "white") +
  geom_point(data = probs_df, aes(x = x, y = y), pch = 19) +
  geom_label(data = prob_labels_df, aes(x = x - 2*width, y = y, label = label, color = region), parse = TRUE, hjust = 0) +
  labs(x = "Input to Inverse-Logit Function", y = "Probability") +
  scale_fill_manual(values = colors_J) + 
  scale_color_manual(values = colors_J) + 
  theme_minimal() +
  theme(legend.position = "none")
```

## Example

As an example, we can model party ID as a function of age and race.

### `polr()`

```{r}
# load data
df <- forcats::gss_cat |>
  mutate(race = fct_drop(race)) |>
  mutate(partyid = case_when(partyid == "No answer" ~ NA,
                             partyid == "Don't know" ~ NA,
                             partyid == "Other party" ~ NA, 
                             TRUE ~ partyid),
         partyid = fct_drop(partyid)) |>
  glimpse()

# look at outcome
levels(df$partyid)

# fit ordered logit model
fit <- MASS::polr(partyid ~ age*race, data = df, Hess = TRUE)

# summarize fit
summary(fit)
```

### Probabilities with \{marginaleffects\}

```{r}
# use {marginaleffects}
library(marginaleffects)
p <- predictions(fit, 
            variables = list(age = 25:75), 
            newdata = datagrid(race = unique)) |>
  glimpse()
ggplot(p, aes(x = age, y = estimate, color = group)) + 
  geom_line() + 
  facet_wrap(vars(race))
```

### Computing cumulative probabilities



```{r}
cumulative_p <- p %>%
  arrange(race, age, group) %>%
  group_by(race, age) %>%
  mutate(cumulative_prob = cumsum(estimate),
         j = as.integer(group),
         cdf_label = paste0("Pr(Y ≤ ", group, ")"),
         cdf_label = reorder(cdf_label, -j)) %>%
  ungroup()
ggplot(cumulative_p, aes(x = age, y = cumulative_prob, color = cdf_label)) + 
  geom_line() + 
  facet_wrap(vars(race)) 
```

We can use `position = "stack"` to trick `ggplot()` into plotting the cumulative probabilities for us if we don't want to compute them manually first.

```{r}
ggplot(p, aes(x = age, y = estimate, color = group)) +
  geom_line(position = "stack") +
  facet_wrap(vars(race))

ggplot(p, aes(x = age, y = estimate, fill = group)) +
  geom_area(position = "stack") +
  facet_wrap(vars(race)) 
```