---
fig-asp: 0.75
---

# Ordered Logit

## Ordered Outcomes

Suppose an ordered outcome $y_i \in \{1, 2, \ldots, J\}$, where $\{1, 2, \ldots, J\}$ correspond to *qualitative labels* that have a natural ordering.

*Example 1.* The ANES asks respondents to place themselves along a 7-point **ideology** scale with positions labeled "extremely liberal," "liberal," "slightly liberal," "moderate/middle of the road," "slightly conservative," "conservative," and "extremely conservative." We typically let 1, 2, 3, 4, 5, 6, and 7 correspond to these seven ordered labels. 

*Example 2*. The **Political Terror Scale** (PTS) is a country-year measure of state repression. This measure assigns one of five qualitative labels summarizing the amount of state repression to each country-year based on reports from Amnesty International and the U.S. State Department.

| Numeric Coding      | Qualitative Label |
|------------|----------------|
| 1  | Countries under a secure rule of law, people are not imprisoned for their views, and torture is rare or exceptional. Political murders are extremely rare. |
| 2  | There is a limited amount of imprisonment for nonviolent political activity. However, few persons are affected, torture and beatings are exceptional. Political murder is rare. |
| 3  | There is extensive political imprisonment, or a recent history of such imprisonment. Execution or other political murders and brutality may be common. Unlimited detention, with or without a trial, for political views is accepted. |
| 4  | Civil and political rights violations have expanded to large numbers of the population. Murders, disappearances, and torture are a common part of life. In spite of its generality, on this level terror affects those who interest themselves in politics or ideas. |
| 5  | Terror has expanded to the whole population. The leaders of these societies place no limits on the means or thoroughness with which they pursue personal or ideological goals. |

## Ordered Logit

Let linear predictor $\eta_i$ be our usual linear predictor so that $\eta_i = X_i\beta$, except *without an intercept $\beta_0$.*

Now define $J-1$ "cutpoints" $\alpha_1 < \alpha_2 < \cdots < \alpha_{J-1}$ that act as a separate intercept for each cumulative probability $\Pr(Y_i \leq j)$.

The **ordered logit** defines the *cumulative* probabilities as

$$
\Pr(Y_i \leq j) = \operatorname{logit}^{-1}(\alpha_j - \eta_i),
\qquad j = 1, \ldots, J-1.
$$

Because the same $\beta$ appears for every $j$, the ordered logit is also called the **proportional odds** model. The log-odds of falling in category $j$ or below is $\log \left[ \dfrac{\Pr(Y_i \leq j)}{1 - \Pr(Y_i \leq j)} \right] = \alpha_j - \eta_i$, so a one-unit increase in $x_{ik}$ changes the log-odds of $Y_i \leq j$ by $-\beta_k$ (multiplies the odds by $e^{-\beta_k}$) for *every* $j$. Because of the minus sign, a positive $\beta_k$ makes the lower categories less likely and the higher categories more likely. This is the parameterization that `MASS::polr()` (which calls the cutpoints `zeta` and prints them as "Intercepts") and the `cumulative()` family in {brms} use.

Then probabilities for each category are

$$
\begin{align}
\Pr(Y_i = 1) &= \operatorname{logit}^{-1}(\alpha_1 - \eta_i) \\
\Pr(Y_i = 2) &= \operatorname{logit}^{-1}(\alpha_2 - \eta_i) - \operatorname{logit}^{-1}(\alpha_{1} - \eta_i) \\
\Pr(Y_i = 3) &= \operatorname{logit}^{-1}(\alpha_3 - \eta_i) - \operatorname{logit}^{-1}(\alpha_{2} - \eta_i) \\
             &\vdots\\
\Pr(Y_i = j) &= \overbrace{\operatorname{logit}^{-1}(\alpha_j - \eta_i)}^{\Pr(Y_i \leq j)} - \overbrace{\operatorname{logit}^{-1}(\alpha_{j-1} - \eta_i)}^{^{\Pr(Y_i \leq j-1)}}, \qquad j = 2, \ldots, J-1 \\
             &\vdots\\
\Pr(Y_i = J) &= 1 - \operatorname{logit}^{-1}(\alpha_{J-1} - \eta_i).
\end{align}
$$

The figure below shows this graphically for an outcome with four possible values.

- I set the three cutpoints to $\alpha = (-2.5,-1.3,1.3)$. 
- I set the linear predictor for an example observation to $\eta_i = -1.5$. 

The figure shows how $\alpha$ and $\eta_i$ are translated into $\Pr(Y_i = j)$.

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-asp: 0.75
#| fig-width: 12
library(tidyverse)
library(latex2exp)


# parameters
alpha <- sort(c(-2.5, -1.3, 1.3))
J <- length(alpha) + 1
eta   <- -1.5   

# make a color pallette
library(RColorBrewer)
colors_J <- brewer.pal(J, "Dark2")  # or "Set1", "Paired", "Accent"

cumulative_prob <- c(plogis(alpha - eta), 1)
prob <- c(cumulative_prob[1], diff(cumulative_prob))

# S-shaped curve
curve_df <- tibble(x = seq(-8, 8, by = 0.001)) |>
  mutate(prob = plogis(x))

# lines
x_lines_df <- tibble(x0 = alpha - eta, x1 = alpha - eta, 
                     y0 = 0, y1 = cumulative_prob[1:(J - 1)])

y_lines_df <- tibble(x0 = min(curve_df$x), x1 = alpha - eta, 
                     y0 = cumulative_prob[1:(J - 1)], y1 = cumulative_prob[1:(J - 1)])

# labels
x_labs <- paste0(
  r"($\alpha_{)", 1:(J - 1), r"(} - \eta_i = )",
  scales::number(alpha - eta, accuracy = 0.01),
  r"($)"
)
x_labels_df <- tibble(
  x = alpha - eta,
  y = cumulative_prob[1:(J - 1)]/2,
  label = TeX(x_labs)
)
y_labs <- paste0(
  r"($\Pr(Y_i \leq )", 1:(J - 1), ")",
  r"(=logit^{-1}(\alpha_{)", 1:(J - 1),
  r"(} - \eta_i)=)",
  scales::number(cumulative_prob[1:(J - 1)], accuracy = 0.01),
  r"($)"
)
y_labels_df <- tibble(
  x = (min(curve_df$x) + (alpha - eta)) / 2,
  y = cumulative_prob[1:(J - 1)],
  label = TeX(y_labs), 
  region = as.character(1:(J-1))
)

# probability labels
probs_df <- tibble(x = min(curve_df$x), y = cumulative_prob)
prob_labs <- paste0(
  r"($\Pr(Y_i = )", 1:J, ") = ",
  r"(\Pr(Y_i \leq )", (1:J), ") - ",
  r"(\Pr(Y_i \leq )", 1:J - 1, ") = ", 
  scales::number(prob[1:J], accuracy = 0.01),
  r"($)"
)
prob_labs[1] <- paste0(
  r"($\Pr(Y_i = 1) = \Pr(Y_i \leq 1) = )", 
  scales::number(prob[1], accuracy = 0.01),
  r"($)"
)
prob_labs[J] <- paste0(
  r"($\Pr(Y_i = )", J, ") =", r"(1 - \Pr(Y_i \leq )", J - 1, ") =", 
  scales::number(prob[J], accuracy = 0.01),
  r"($)"
)
prob_labels_df <- tibble(
  x = min(curve_df$x),
  y = cumulative_prob - diff(c(0, cumulative_prob))/2,
  label = TeX(prob_labs), 
  region = as.character(1:(J))
)

# y polygons: use a for-loop to build the polygons one at a time
y_poly_list <- NULL
y_poly_list[[1]] <- curve_df |>
  filter(x <= alpha[1] - eta) |>
  mutate(to = cumulative_prob[1], 
         from = prob) |>
  mutate(region = as.character(1))
for (j in 2:(J - 1)) {
  y_poly_list[[j]] <- curve_df |>
    filter(x <= alpha[j] - eta) |>
    mutate(to = cumulative_prob[j], 
           from = ifelse(cumulative_prob[j-1] > prob, cumulative_prob[j - 1], prob)) |>
    mutate(region = as.character(j))
}
y_poly_list[[J]] <- curve_df |>
  mutate(to = 1, 
         from = ifelse(cumulative_prob[J-1] > prob, cumulative_prob[J - 1], prob)) |>
  mutate(region = as.character(J))
y_poly_df <- bind_rows(y_poly_list)

# x polygons: use a for-loop to build the polygons one at a time
x_poly_list <- NULL
x_poly_list[[1]] <- curve_df |>
  filter(x <= alpha[1] - eta) |>
  mutate(to = prob, 
         from = 0) |>
  mutate(region = as.character(1))
for (j in 2:(J - 1)) {
  x_poly_list[[j]] <- curve_df |>
    filter(x <= alpha[j] - eta, x > alpha[j - 1] - eta) |>
    mutate(to = prob, 
           from = 0) |>
    mutate(region = as.character(j))
}
x_poly_list[[J]] <- curve_df |>
  filter(x > alpha[J - 1] - eta) |>
  mutate(to = prob, 
         from = 0) |>
  mutate(region = as.character(J))
x_poly_df <- bind_rows(x_poly_list)

# probability blocks
width <- diff(range(curve_df$x))*0.05
block_list <- NULL
block_list[[1]] <- tibble(x = seq(min(curve_df$x) - width, min(curve_df$x), by = 0.01)) |>
  mutate(from = 0, to = cumulative_prob[1], region = "1") 
for (j in 2:(J - 1)) {
  block_list[[j]] <- tibble(x = seq(min(curve_df$x) - width, min(curve_df$x), by = 0.01)) |>
    mutate(from = cumulative_prob[j-1], to = cumulative_prob[j], region = as.character(j))
}
block_list[[J]] <- tibble(x = seq(min(curve_df$x) - width, min(curve_df$x), by = 0.01)) |>
  mutate(from = cumulative_prob[J - 1], to = cumulative_prob[J], region = as.character(J))


block_df <- bind_rows(block_list)

ggplot() +
  geom_ribbon(data = block_df, aes(x = x, ymin = from, ymax = to, fill = region), alpha = 1.0, color = "black") +
  geom_segment(aes(x = min(curve_df$x), y = 0, yend = 1), color = "black") + 
  geom_segment(aes(x = min(curve_df$x) - width, y = 0, yend = 1), color = "black") + 
  geom_ribbon(data = x_poly_df, aes(x = x, ymin = from, ymax = to, fill = region), alpha = 0.35) +
  geom_ribbon(data = y_poly_df, aes(x = x, ymin = from, ymax = to, fill = region), alpha = 0.35) +
  
  geom_segment(data = x_lines_df, aes(x = x0, xend = x1, y = y0, yend = y1), linetype = "dotted") +
  geom_label(data = x_labels_df, aes(x = x, y = y, label = label), parse = TRUE) +
  geom_segment(data = y_lines_df, aes(x = x0, xend = x1, y = y0, yend = y1), linetype = "dotted") +
  geom_segment(data = y_labels_df, aes(x = x, xend = x, y = y, yend = 0, color = region), linewidth = 4) + 
  
  geom_label(data = y_labels_df, aes(x = x, y = y, label = label, color = region), parse = TRUE) +
  geom_line(data = curve_df, aes(x = x, y = prob)) +
  geom_point(data = x_lines_df, aes(x = x1, y = y1), shape = 21, fill = "white") +
  geom_point(data = probs_df, aes(x = x, y = y), pch = 19) +
  geom_label(data = prob_labels_df, aes(x = x - 2*width, y = y, label = label, color = region), parse = TRUE, hjust = 0) +
  labs(x = "Input to Inverse-Logit Function", y = "Probability") +
  scale_fill_manual(values = colors_J) + 
  scale_color_manual(values = colors_J) + 
  theme_minimal() +
  theme(legend.position = "none")
```

## Example: Approval of President Bush, 1992

- Data: 1992 ANES Time Series Study [@anes1992], as cleaned by Chris Adolph for his maximum likelihood course; the model and the codebook are his Problem Set 4 [@adolph2025].
- Outcome: approval of President Bush's job performance, four categories.
- Predictors: opposition to using military force (1–5), distance between the respondent's ideology and Bush's (0–6), evaluation of the economy over the past year (1 = much better … 5 = much worse), party ID (−3 strong Democrat … 3 strong Republican), years of education.
- Ordered *probit*, as in the problem set: `method = "probit"`.
- 750 respondents; 558 complete on these variables.

```{r}
library(tidyverse)
library(MASS)  # masks dplyr::select()
library(marginaleffects)

# load data
approval <- read_csv("data/bush-approval-1992.csv") |>
  mutate(bush_approval = factor(bush_approval, ordered = TRUE)) |>
  glimpse()

# check the order of the outcome
levels(approval$bush_approval)

# fit ordered probit
fit_approval <- polr(bush_approval ~ military_force + ideology_distance +
                       economy + party_id + education,
                     data = approval, method = "probit", Hess = TRUE)
summary(fit_approval)
```

- Positive `party_id` coefficient: higher categories more likely (sign convention above), so Republicans approve more.
- Negative coefficients: opposing military force, ideological distance from Bush, and a worse economy each move respondents toward disapproval.
- Cutpoints on the probit scale; not directly interpretable.

```{r}
#| fig-asp: 0.5
# probabilities across military_force for strong Democrats and strong Republicans
p <- predictions(fit_approval,
                 newdata = datagrid(military_force = 1:5,
                                    party_id = c(-3, 3))) |>
  glimpse()

ggplot(p, aes(x = military_force, y = estimate, fill = group)) +
  geom_area(position = "stack") +
  facet_wrap(vars(party_id), labeller = label_both)
```

- Other predictors at their means (`datagrid()` default).
- Strong Democrats: Pr(strongly disapprove) about 0.28 at `military_force = 1`, about 0.77 at 5.
- Strong Republicans: Pr(strongly approve) about 0.49 at 1, about 0.09 at 5.

### Computing cumulative probabilities {#cumulative-probabilities}

- The model is defined through $\Pr(Y_i \leq j)$; `predictions()` returns $\Pr(Y_i = j)$, one row per category.

```{r}
# cumulative probabilities: Pr(Y <= j) and Pr(Y >= j)
cumulative_p <- p |>
  arrange(party_id, military_force, group) |>                # <1>
  group_by(party_id, military_force) |>                      # <2>
  mutate(pr_le = cumsum(estimate),                           # <3>
         pr_ge = 1 - lag(cumsum(estimate), default = 0)) |>  # <4>
  ungroup()

# one grid point: strong Democrats at military_force = 3
cumulative_p |>
  filter(party_id == -3, military_force == 3) |>
  dplyr::select(group, estimate, pr_le, pr_ge)
```
1. Sort the categories into their order within each grid point, so the running sum accumulates from category 1 upward.
2. One running sum per grid point, i.e., per combination of `party_id` and `military_force`.
3. $\Pr(Y_i \leq j)$: the running sum $\Pr(Y_i = 1) + \cdots + \Pr(Y_i = j)$.
4. $\Pr(Y_i \geq j) = 1 - \Pr(Y_i \leq j - 1)$: `lag()` shifts the running sum down one category, and `default = 0` covers $j = 1$.

- $\Pr(Y_i \leq J)$ and $\Pr(Y_i \geq 1)$ are always 1; drop them before plotting.

```{r}
#| fig-asp: 0.5
# plot Pr(Y <= j) for j = 1, 2, 3
cumulative_p |>
  filter(group != "4. Strongly approve") |>
  ggplot(aes(x = military_force, y = pr_le, color = group)) +
  geom_line() +
  geom_point() +
  facet_wrap(vars(party_id), labeller = label_both) +
  labs(y = "Pr(Y <= j)")
```

- Strong Democrats at `military_force = 3`: Pr(strongly disapprove) 0.54, Pr(disapprove or worse) 0.83, Pr(approve or worse) 0.99.
- Each curve is one cutpoint: $\Pr(Y_i \leq j) = \Phi(\alpha_j - \eta_i)$ here (probit), so the three curves move together as $\eta_i$ changes. That is the proportional-odds (parallel) structure.
- The stacked `geom_area()` above draws the same curves, but *upper-tail*: `position = "stack"` puts the first level on top, so the boundary above band $j$ is $\Pr(Y_i \geq j)$, not $\Pr(Y_i \leq j)$. At `military_force = 3` for strong Democrats the boundaries sit at 0.46, 0.17, and 0.01, the `pr_ge` column. `position_stack(reverse = TRUE)` puts level 1 at the bottom and makes the boundaries $\Pr(Y_i \leq j)$.

## Example: Working mothers, GSS 1977 and 1989

- Data: General Social Survey, 1977 and 1989 [@gss2019]; the `ordwarm2` data of @long1997 [ch. 5] and @longfreese2014 [ch. 7].
- Outcome: agreement with "A working mother can establish just as warm and secure a relationship with her children as a mother who does not work," four categories.
- Predictors: survey year, gender, race, age, years of education, occupational prestige.
- Ordered logit, Long & Freese's specification.
- 2,293 respondents, no missing values.

```{r}
library(tidyverse)
library(MASS)  # masks dplyr::select()
library(marginaleffects)

# load data
warm <- read_csv("data/working-mothers-gss.csv") |>
  mutate(warm = factor(warm, ordered = TRUE)) |>
  glimpse()

# check the order of the outcome
levels(warm$warm)

# fit ordered logit
fit_warm <- polr(warm ~ factor(year) + gender + race + age + education + prestige,
                 data = warm, Hess = TRUE)
summary(fit_warm)
```

- Matches Long & Freese: `factor(year)1989` 0.52, `genderMale` −0.73, `raceWhite` −0.39, `age` −0.02, `education` 0.07, `prestige` 0.006.
- Men, whites, and older respondents less likely to agree; 1989 respondents and the more educated more likely.

```{r}
#| fig-asp: 0.5
# probabilities across age, by gender
p <- predictions(fit_warm,
                 newdata = datagrid(age = 18:89,
                                    gender = unique)) |>
  glimpse()

ggplot(p, aes(x = age, y = estimate, fill = group)) +
  geom_area(position = "stack") +
  facet_wrap(vars(gender))
```

- Year and race at their modes, education and prestige at their means.
- Women: Pr(strongly agree) about 0.26 at age 20, about 0.09 at 80.
- Men: Pr(strongly disagree) about 0.12 at 20, about 0.34 at 80.
- Band boundaries are $\Pr(Y_i \geq j)$ (first level stacked on top).

```{r}
#| fig-asp: 0.5
# cumulative probabilities Pr(Y <= j), then plot j = 1, 2, 3
cumulative_p <- p |>
  arrange(gender, age, group) |>
  group_by(gender, age) |>
  mutate(pr_le = cumsum(estimate)) |>
  ungroup()

cumulative_p |>
  filter(group != "4. Strongly agree") |>
  ggplot(aes(x = age, y = pr_le, color = group)) +
  geom_line() +
  facet_wrap(vars(gender)) +
  labs(y = "Pr(Y <= j)")
```

- Three parallel curves per panel, one per cutpoint; men's curves sit above women's (less agreement).

## Example: Political ideology by party and gender

- Data: Table 6.7 of @agresti2007 [sec. 6.2.2], a 2 × 2 × 5 table from the General Social Survey (no year given), expanded to one row per respondent (N = 835); the party-only version is the UVA StatLab tutorial [@ford2015].
- Outcome: political ideology, five categories from very liberal to very conservative.
- Predictors: party (Democrat, Republican) and gender.
- Ordered logit; Agresti's cumulative logit (proportional odds) model.

```{r}
library(tidyverse)
library(MASS)  # masks dplyr::select()
library(marginaleffects)

# load data
ideology <- read_csv("data/ideology-party-gender.csv") |>
  mutate(ideology = factor(ideology, ordered = TRUE)) |>
  glimpse()

# check the order of the outcome
levels(ideology$ideology)

# fit ordered logit
fit_ideology <- polr(ideology ~ party + gender,
                     data = ideology, Hess = TRUE)
summary(fit_ideology)
```

- Agresti fits party alone and reports 0.975 (SE 0.129) [@agresti2007, p. 183]; with gender added, `partyRepublican` 0.96 (SE 0.13) and `genderMale` 0.12 (SE 0.13).
- Republicans much more conservative; gender difference small and uncertain.

```{r}
#| fig-asp: 0.5
# probabilities for the four groups
p <- predictions(fit_ideology,
                 newdata = datagrid(party = unique,
                                    gender = unique)) |>
  glimpse()

ggplot(p, aes(x = party, y = estimate, fill = group)) +
  geom_col() +
  facet_wrap(vars(gender))
```

- Both predictors binary, so `x` takes two values. `geom_area()` needs a numeric `x` and draws nothing for a character `x`; `geom_col()` stacks the same probabilities as columns.
- Women: Pr(very conservative) about 0.11 for Democrats, about 0.25 for Republicans.
- Column boundaries are $\Pr(Y_i \geq j)$ (first level stacked on top), as in the two figures above.

```{r}
#| fig-asp: 0.5
# cumulative probabilities Pr(Y <= j), then plot j = 1, ..., 4
cumulative_p <- p |>
  arrange(gender, party, group) |>
  group_by(gender, party) |>
  mutate(pr_le = cumsum(estimate)) |>
  ungroup()

cumulative_p |>
  filter(group != "5. Very conservative") |>
  ggplot(aes(x = party, y = pr_le, color = group, group = group)) +
  geom_line() +
  geom_point() +
  facet_wrap(vars(gender)) +
  labs(y = "Pr(Y <= j)")
```

- Every cumulative probability drops from Democrats to Republicans by the same log-odds shift, 0.96.
