# Integration

## Definition

If derivatives describe the rate of change of a function, integrals do the opposite—they **accumulate** values over an interval.

::: {#def-integral}
The **definite integral** of a function $f(x)$ from $a$ to $b$ is defined as:

$$
\int_a^b f(x)\, dx = \lim_{n \to \infty} \sum_{i=1}^n f(x_i^*) \Delta x
$$

where the interval $[a, b]$ is divided into $n$ subintervals of width $\Delta x = \frac{b - a}{n}$ and $x_i^*$ is a sample point in the $i$th subinterval.
:::

This limit represents the total area under the curve $f(x)$ from $a$ to $b$.

### Indefinite Integral

An **indefinite integral** (or **antiderivative**) is a function whose derivative is the original function.

::: {#def-antideriv}
If $F'(x) = f(x)$, then $F(x)$ is an **antiderivative** of $f(x)$, and we write:

$$
\int f(x)\, dx = F(x) + C
$$

where $C$ is an arbitrary constant.
:::

---

## Physical Interpretation of Integrals

Suppose your car is moving at a velocity $f(x)$, where $x$ is time and $f(x)$ is in meters per second. Then:

- $\int_a^b f(x)\, dx$ is the **total distance traveled** between time $x = a$ and $x = b$.

In this way, **integration undoes differentiation**. If the derivative of position is velocity, then the integral of velocity is position.

| Quantity     | Symbol            | Interpretation                        | Units (if $x$ is time)         |
|--------------|-------------------|----------------------------------------|---------------------------------|
| Velocity     | $f(x)$            | Rate of change of position             | Meters per second (m/s)         |
| Position     | $\int f(x)\, dx$  | Total distance traveled (accumulated) | Meters (m)                      |

The key idea is this: **integration accumulates change**. It’s the natural inverse of differentiation.

---

## Why This Matters

Many real-world questions ask about **totals** or **areas**:

- **How much profit was earned over the year?**
- **What is the total rainfall over a week?**
- **What’s the area under the likelihood curve?**
- **How much change accumulates over time?**

All of these questions require integration. Let’s build some fluency with it.

---

## Basic Rules of Integration

### Constant Rule

::: {#thm-int-const}
## Constant Rule
If $f(x) = a$, then:

$$
\int a\, dx = ax + C
$$
:::

::: {#exm-int1}
Compute $\int 5\, dx$.
:::

*Solution.* Use the constant rule: $\int 5\, dx = 5x + C$.

---

### Power Rule

::: {#thm-int-power}
## Power Rule
If $f(x) = x^n$, then:

$$
\int x^n\, dx = \frac{x^{n + 1}}{n + 1} + C,\quad \text{for } n \ne -1
$$
:::

::: {#exm-int2a}
Compute $\int x^3\, dx$.
:::

*Solution.* Apply the power rule:

$$
\int x^3\, dx = \frac{x^4}{4} + C
$$

::: {#exm-int2b}
Compute $\int x^5 - 2x^2 + 7\, dx$.
:::

*Solution.* Integrate each term using the power rule:

$$
\int x^5\, dx = \frac{x^6}{6},\quad
\int -2x^2\, dx = -\frac{2x^3}{3},\quad
\int 7\, dx = 7x
$$

So the full result is:

$$
\frac{x^6}{6} - \frac{2x^3}{3} + 7x + C
$$

---

### Exponential Rule

::: {#thm-int-exp}
## Exponential Rule
If $f(x) = e^x$, then:

$$
\int e^x\, dx = e^x + C
$$
:::

---

### Logarithm Rule

::: {#thm-int-log}
## Logarithm Rule
If $f(x) = \frac{1}{x}$, then:

$$
\int \frac{1}{x}\, dx = \log|x| + C,\quad x \ne 0
$$
:::

---

### Sum Rule

::: {#thm-int-sum}
## Sum Rule
If $f(x) = g(x) + h(x)$, then:

$$
\int f(x)\, dx = \int g(x)\, dx + \int h(x)\, dx
$$
:::

::: {#exm-int3}
Compute $\int (x^2 + \frac{1}{x})\, dx$.
:::

*Solution.* Apply the sum and power rules:

$$
\int x^2\, dx = \frac{x^3}{3},\quad
\int \frac{1}{x}\, dx = \log|x|
$$

So:

$$
\int (x^2 + \frac{1}{x})\, dx = \frac{x^3}{3} + \log|x| + C
$$

---

### Integration by Substitution (Chain Rule Reversed)

Substitution allows us to integrate composite functions, reversing the chain rule.

::: {#thm-int-sub}
## Substitution Rule
Let $u = g(x)$. Then:

$$
\int f(g(x)) g'(x)\, dx = \int f(u)\, du
$$
:::

::: {#exm-int4}
Compute $\int 2x \cdot \exp(x^2)\, dx$.
:::

*Solution.* Let $u = x^2$, so $du = 2x\, dx$. Then:

$$
\int 2x \cdot \exp(x^2)\, dx = \int \exp(u)\, du = \exp(u) + C = \exp(x^2) + C
$$

---

### Integration by Parts (Product Rule Reversed)

Sometimes, it helps to reverse the product rule.

::: {#thm-int-parts}
## Integration by Parts
If $u = u(x)$ and $v = v(x)$, then:

$$
\int u\, dv = uv - \int v\, du
$$
:::

::: {#exm-int5}
Compute $\int x \cdot \log(x)\, dx$.
:::

*Solution.* Use integration by parts:

Let $u = \log(x)$, so $du = \frac{1}{x} dx$  
Let $dv = x\, dx$, so $v = \frac{x^2}{2}$

Then:

$$
\int x \log(x)\, dx = \frac{x^2}{2} \log(x) - \int \frac{x^2}{2} \cdot \frac{1}{x}\, dx
= \frac{x^2}{2} \log(x) - \int \frac{x}{2}\, dx
= \frac{x^2}{2} \log(x) - \frac{x^2}{4} + C
$$

---

## Definite Integrals

If we want to compute the **total accumulated value** over a specific interval $[a, b]$, we use definite integrals.

::: {#def-definite}
The definite integral from $a$ to $b$ is:

$$
\int_a^b f(x)\, dx = F(b) - F(a)
$$

where $F(x)$ is any antiderivative of $f(x)$.
:::

::: {#exm-def1}
Compute $\int_1^2 x^2\, dx$.
:::

*Solution.* Find the antiderivative: $F(x) = \frac{x^3}{3}$

Evaluate:

$$
\int_1^2 x^2\, dx = \frac{2^3}{3} - \frac{1^3}{3} = \frac{8 - 1}{3} = \frac{7}{3}
$$

---

## Integration in R

R does not compute symbolic integrals with base functions, but numerical integration is straightforward using `integrate()`.

::: {#exm-rint1}
Compute $\int_1^2 x^2\, dx$ numerically.
:::

```{r}
f <- function(x) x^2
integrate(f, lower = 1, upper = 2)
```

Returns:

```
7/3 = 2.333...
```

::: {#exm-rint2}
Compute $\int_0^1 x \log(x)\, dx$ numerically.
:::

```{r}
f <- function(x) ifelse(x == 0, 0, x * log(x))
integrate(f, 0, 1)
```

This returns approximately:

```
-0.25
```

---

Numerical integration is useful for functions that have no closed-form antiderivative or are only defined computationally.

---

## Summary: Derivatives vs Integrals

| Operation     | Symbol                      | Meaning                         |
|---------------|-----------------------------|---------------------------------|
| Derivative    | $f'(x)$ or $\frac{df}{dx}$  | Instantaneous rate of change    |
| Integral      | $\int f(x)\, dx$            | Accumulated change (area under curve) |
| Definite Int. | $\int_a^b f(x)\, dx$        | Total change from $x = a$ to $x = b$ |

Together, derivatives and integrals are the fundamental tools of calculus. They describe change and accumulation—central ideas in modeling, statistics, economics, physics, and beyond.
