# Fractions, Exponents, and Logarithms

## Fractions

You can manipulate fractions with the identities below:

- $\dfrac{a}{c} + \dfrac{b}{c} = \dfrac{a + b}{c}$

- $\dfrac{a}{c} - \dfrac{b}{c} = \dfrac{a - b}{c}$

- $\dfrac{a}{b} \times \dfrac{c}{d} = \dfrac{a \times c}{b \times d}$

- $\dfrac{\frac{a}{b}}{\frac{c}{d}} = \dfrac{a}{b} \times \dfrac{d}{c} = \dfrac{a \times d}{b \times c}$

::: {.callout-warning}
**Potential Pitfalls**

- Do **not** add fractions with different denominators directly. For example, $\frac{1}{2} + \frac{1}{3} \ne \frac{2}{5}$ and $\frac{1}{a} + \frac{1}{b} \ne \frac{1}{a + b}$.
:::

::: {#exm-fraction1}
Compute $\frac{1}{4} + \frac{3}{4}$.
:::

*Solution.* The denominators match, so add the numerators to obtain $\frac{1 + 3}{4} = \frac{4}{4} = 1$.

::: {#exm-fraction2}
Compute $\frac{2}{3} \times \frac{5}{7}$.
:::

*Solution.* Multiply numerators and denominators to obtain $\frac{2 \times 5}{3 \times 7} = \frac{10}{21}$.

::: {#exm-fraction3}
Compute $\dfrac{\frac{5}{6}}{\frac{2}{9}}$, which is the same as ${\frac{5}{6}} \div {\frac{2}{9}}$.
:::

*Solution.* Multiply by the reciprocal to obtain $\frac{5}{6} \times \frac{9}{2} = \frac{5 \times 9}{6 \times 2} = \frac{45}{12} = \frac{15}{4}$.

::: {#exm-fraction4}
Simplify $\frac{x}{3} + \frac{2x}{3}$.
:::

*Solution.* The denominator matches, so add the numerators to obtain $\frac{x + 2x}{3} = \frac{3x}{3} = x$.

::: {#exm-fraction5}
Simplify $\frac{\alpha}{\beta} \times \frac{\gamma}{\alpha}$.
:::

*Solution.* Multiply across to obtain $\frac{\alpha \times \gamma}{\beta \times \alpha}$. Cancel $\alpha$ from numerator and denominator to obtain $\frac{\gamma}{\beta}$.

::: {#exm-fraction6}
Simplify $\frac{y_1}{n} + \frac{y_2 + y_3}{n}$.
:::

*Solution.* The denominator matches, so add the numerators to obtain $\frac{y_1 + y_2 + y_3}{n}$.

## Exponents and Logarithms

Exponents and logarithms are inverse operations. If $b^x = y$, then $\log_b(y) = x$. In this course, we mostly use the natural logarithm, which uses base $e \approx 2.718$. We denote the natural log as $\log(x)$ (not as $\ln(x)$). In plots, we'll sometimes use the base-10 exponents and logarithms $10^x$ and $\log_{10}(x)$.

You can manipulate exponents and logarithms with the identities below:

- $e^{\log(x)} = x$ for $x > 0$
- $\log(e^x) = x$
- $\log(ab) = \log(a) + \log(b)$
- $\log\left(\dfrac{a}{b}\right) = \log(a) - \log(b)$
- $\log(a^b) = b\log(a)$
- $a^{\log(b)} = b^{\log(a)}$

::: {.callout-warning}
## Common Pitfalls

- $\log(a + b) \ne \log(a) + \log(b)$
- $\log(a - b) \ne \log(a) - \log(b)$
- $\log(x)$ is undefined for $x \le 0$
:::

::: {#exm-log-1}
Evaluate $\log(e^3)$.
:::

*Solution.* $\log(e^3) = 3 \log(e) = 3$.

::: {#exm-exp-1}
Evaluate $e^{\log(7)}$.
:::

*Solution.* Since exponentiation and log are inverse operations, so $e^{\log(7)} = 7$. In general, $e^{\log(x)} = x$.

::: {#exm-log-2}
Simplify then evaluate $\log(8) + \log(2)$.
:::

*Solution.* $\log(8 \times 2) = \log(16)$. Using R, $\log(16) \approx 2.7726$.

```{r}
log(16)
```

::: {#exm-log-3}
Simplify $\log(a^3b^2)$.
:::

*Solution.* $\log(a^3b^2) = \log(a^3) + \log(b^2) = 3\log(a) + 2\log(b)$

::: {#exm-log-4}
Simplify $\log\left(\dfrac{x^4}{y^2}\right)$.
:::

*Solution.* $\log\left(\dfrac{x^4}{y^2}\right) = \log(x^4) - \log(y^2) = 4\log(x) - 2\log(y)$

::: {#exm-log-5}
Simplify $\log\left(\prod_{i=1}^n y_i\right)$.
:::

*Solution.* $\log\left(\prod_{i=1}^n y_i\right) = \sum_{i=1}^n \log(y_i)$. This follows from the identity $\log(ab) = \log(a) + \log(b)$. See the example below in R.

```{r}
# example
y <- 1:5
log(prod(y))
sum(log(y))
```


## Solving Equations

To **solve** an equation, find all values of the variable(s) that make the equation true.

::: {#exm-solve-1}
Solve $\frac{x}{2} + 3 = 5$.
:::

*Solution.* Subtract 3 from both sides to obtain $\frac{x}{2} = 2$. Multiply both sides by 2 to obtain $x = 4$.

::: {#exm-solve-2}
Solve $\frac{2y - 1}{3} = 3$.
:::

*Solution.* Multiply both sides by 3 to obtain $2y - 1 = 9$. Add 1 to obtain $2y = 10$. Divide by 2 to obtain $y = 5$.

::: {#exm-solve-3}
Solve $x^2 = 4$.
:::

*Solution.* Take square roots to obtain $x = \pm 2$. Notice that there are two solutions. The plot below shows that the curve $x^2 - 4$ crosses the $x$-axis in two places (i.e., at $x = \pm 2$, which are the solutions we found above).

```{r}
# load packages
library(ggplot2)

# create function
f <- function(x) x^2 - 4

# plot the function to see the two solutions
ggplot() +
  xlim(-3, 3) + 
  stat_function(fun = f) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(y = "f(x)", 
       x = "x")
```


::: {#exm-solve-4}
Solve $e^x = 7$.
:::

*Solution.* Take natural logs to obtain $x = \log(7)$. Using R, $\log(y) \approx 1.9459$.

```{r}
log(7)
```

::: {#exm-solve-5}
Solve $\log(x) = 2$.
:::

*Solution.* Exponentiate both sides to obtain $x = e^2$. Using R, $e^2 \approx 7.3891$.

```{r}
exp(2)
```

::: {#exm-solve-6}
Solve $2^x = 16$.
:::

*Solution.* Recognize that $2^4 = 16$, so $x = 4$. 

Alternatively, use logs to obtain 

\begin{align}
\log(2^x) &= \log(16) \\
x \log(2) &= \log(16) \\
x &= \frac{\log(16)}{\log(2)}.
\end{align}

Using R, we can see that $\frac{\log(16)}{\log(2)} = 4$.

```{r}
log(16) / log(2)
```

::: {.callout-tip}
## Change-of-Base Formula 

You might have noticed that $\log(16)/\log(2) = 4$ by recalling the **change-of-base formula** $\log_b(a) = \frac{\log(a)}{\log(b)}$. In this case, we have $\frac{\log(16)}{\log(2)} = \log_2(16)$ (i.e., "To what power must I raise 2 to get 16?"). Since $2^4 = 16$, we have $\log_2(16) = 4$. Thus
$\frac{\log(16)}{\log(2)} = 4$. If we had been clever, we wouldn't have needed a calculator.
:::





::: {#exm-solve-7}
Solve $\log\left(\frac{y}{3}\right) = 2$.
:::

*Solution.* Exponentiate both sides to obtain $\frac{y}{3} = e^2$. Multiply both sides by 3 to obtain $y = 3e^2$. Use R to compute $3e^2 \approx 22.167$.

```{r}
3 * exp(2)
```

::: {#exm-solve-9}
Solve $\log(x^2 + 1) = 3$.
:::

*Solution.* Exponentiate both sides to obtain $x^2 + 1 = e^3$. Subtract 1 from both sides to obtain $x^2 = e^3 - 1$. Take the square root to obtain $x = \pm \sqrt{e^3 - 1} \approx \pm 4.369$.

```{r}
sqrt(exp(3) - 1)
```

::: {#exm-solve-10}
Solve $3a + 2 = 5a - 4$.
:::

*Solution.* Subtract $3a$ from both sides to obtain $2 = 2a - 4$. Add 4 to obtain $6 = 2a$. Divide by 2 to obtain $a = 3$.

::: {#exm-solve-11}
Solve $\alpha^2 + 3\alpha = 0$.
:::

*Solution.* Factor the left-hand side to obtain $\alpha(\alpha + 3) = 0$. Then it's clear that $\alpha = 0$ or $\alpha = -3$. The plot below shows the two solutions.

```{r}
# load packages
library(ggplot2)

# create function
f <- function(x) x^2 + 3*x

# plot the function to see the two solutions
ggplot() +
  xlim(-4, 2) + 
  stat_function(fun = f) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(y = "f(x)", 
       x = "x")
```

::: {#exm-solve-12}
Solve $y_1 - 2y_2 = 4$ for $y_1$ in terms of $y_2$.
:::

*Solution.* Add $2y_2$ to both side to obtain $y_1 = 4 + 2y_2$.
