# Matrices

## Definitions

A **scalar** is a single number.

A **vector** is a 1-dimensional array of numbers.^[By default, we treat these as $1 \times n$ *column* vectors (see definition of *matrix* below).]

A **matrix** is a two-dimensional array of numbers.

- $x$ is $n \times 1$ (column vector).
- $x'$ is $1 \times n$ (row vector).
- $A$ is $n \times p$ (matrix).

Matrices allow us to efficiently describe and perform complex calculations involving many, many numbers (e.g., data sets).

::: {#exm-matrix1}
```{r}
x <- c(1, 2, 3)           # 3 x 1 (column vector)
x
t(x)                      # 1 x 3 (row vector); t() finds transpose, see below
A <- matrix(1:6, nrow = 2)
A                         # 2 x 3 matrix
```
:::

## Transpose

The **transpose** of a matrix flips its rows and columns. If $A$ is $n \times p$, then $A'$ is $p \times n$.

Key identity: $(AB)' = B'A'$

::: {#exm-matrix2}
```{r}
A <- matrix(c(1, 2, 3, 4), nrow = 2)

A
t(A)  # transpose

A2 <- matrix(c(1, 2, 3, 4), nrow = 2, byrow = TRUE)  # note the byrow arg
A2


B <- matrix(c(5, 6, 7, 8), nrow = 2)
AB  <- A %*% B            # matrix multiplication
t(AB)                     # transpose of product
t(B) %*% t(A)             # product of transposes
```
:::

## Matrix Multiplication

### Definition and Computation

The **matrix product** $C = AB$ is defined when the number of columns in $A$ equals the number of rows in $B$. If $A$ is an $n \times p$ matrix and $B$ is a $p \times q$ matrix, then the product $C = AB$ is an $n \times q$ matrix.

The $(i, j)$-entry of $C$ is computed as the *dot product* of the $i$th row of $A$ and the $j$th column of $B$:

$$
C_{ij} = \sum_{k=1}^{p} A_{ik} B_{kj}
$$

That is, to compute the entry in the $i$th row and $j$th column of $C$, multiply corresponding elements from the $i$th row of $A$ and the $j$th column of $B$, and then sum the results.

::: {#exm-matrix-mult-check}

Suppose 
$$
A = \begin{bmatrix}
1 & 2 & 3 & 4 \\
5 & 6 & 7 & 8 \\
9 & 0 & 1 & 2
\end{bmatrix}, \quad
B = \begin{bmatrix}
1 & 0 \\
0 & 1 \\
1 & 0 \\
0 & 1
\end{bmatrix}
$$.

The product $C = AB$ is a $3 \times 2$ matrix. Each entry is computed as $C_{ij} = \sum_{k=1}^4 A_{ik} B_{kj}$.

$$
C = AB = \begin{bmatrix}
1 \cdot 1 + 2 \cdot 0 + 3 \cdot 1 + 4 \cdot 0 & 1 \cdot 0 + 2 \cdot 1 + 3 \cdot 0 + 4 \cdot 1 \\
5 \cdot 1 + 6 \cdot 0 + 7 \cdot 1 + 8 \cdot 0 & 5 \cdot 0 + 6 \cdot 1 + 7 \cdot 0 + 8 \cdot 1 \\
9 \cdot 1 + 0 \cdot 0 + 1 \cdot 1 + 2 \cdot 0 & 9 \cdot 0 + 0 \cdot 1 + 1 \cdot 0 + 2 \cdot 1
\end{bmatrix}
$$

Simplifying, we have 

$$
= \begin{bmatrix}
1 + 0 + 3 + 0 & 0 + 2 + 0 + 4 \\
5 + 0 + 7 + 0 & 0 + 6 + 0 + 8 \\
9 + 0 + 1 + 0 & 0 + 0 + 0 + 2
\end{bmatrix}
= \begin{bmatrix}
4 & 6 \\
12 & 14 \\
10 & 2
\end{bmatrix}
$$.

We can confirm our answer with R.

```{r}
A <- matrix(c(
  1, 2, 3, 4,
  5, 6, 7, 8,
  9, 0, 1, 2
), nrow = 3, byrow = TRUE)

B <- matrix(c(
  1, 0,
  0, 1,
  1, 0,
  0, 1
), nrow = 4, byrow = TRUE)

A %*% B
```
:::

::: {#exm-matrix9}
Let $X$ be a $3 \times 2$ matrix and $\beta$ a $2 \times 1$ vector. Then $X\beta$ is a $3 \times 1$ vector. $X'X$ is $2 \times 2$ and symmetric.

We have a couple of familiar matrix multiplications from linear regression. $X\beta$ is especially important to us!.

```{r}
X <- matrix(c(1, 1, 1, 2, 3, 4), nrow = 3)
beta <- c(0.5, 1)
X %*% beta                          # linear prediction
t(X) %*% X                          # 2 x 2 matrix
```
:::

### Rules

- **Associative**: $(AB)C = A(BC)$
- **Distributive**: $A(B + C) = AB + AC$
- **Not commutative**: $AB \neq BA$ in general

::: {#exm-matrix5}
```{r}
A <- matrix(c(1, 2, 3, 4), 2)         # 2 x 2
B <- matrix(c(5, 6, 7, 8), 2)         # 2 x 2
C <- matrix(c(100, 200, 300, 400), 2) # 2 x 2

# same
(A %*% B) %*% C 
A %*% (B %*% C) 

# same
A %*% (B + C) 
(A %*% B) + (A %*% C) 

# different
A %*% B
B %*% A 
```
:::

## Special Matrices

### Identity Matrix

An **identity matrix** is a square matrix with 1s on the diagonal and 0s elsewhere. Denoted $I$.

- $AI = A$
- $IA = A$
- $I_n$ is $n \times n$

```{r}
I <- diag(3)  # shortcut to make 3x3 identity matrix
I
A <- matrix(1:9, nrow = 3)
A %*% I                           # same as A
```

### Diagonal Matrix

A **diagonal matrix** has nonzero entries only on the diagonal. These matrices are often used for variances or weights.

```{r}
d <- c(2, 4, 6)
D <- diag(d)  # shortcut to make diagonal matrix
```

### Symmetric Matrix

A matrix $A$ is **symmetric** if $A' = A$.

- $X'X$ is always symmetric.

```{r}
X <- matrix(c(1, 2, 3, 4, 5, 6), nrow = 3)
t(X) %*% X                        # symmetric
```


::: {#exm-matrix1}

Construct the following matrices:

- Identity matrix $I_2$
- Diagonal matrix with entries $1, 2, 3$
- Symmetric matrix $S = \begin{bmatrix}2 & 1 \\ 1 & 3\end{bmatrix}$

```{r}
I2 <- diag(2)
diag_mat <- diag(1:3)
S <- matrix(c(2, 1, 1, 3), 2)
```

:::

::: {#exm-matrix2}

Let $\Sigma = \begin{bmatrix}4 & 0 & 0 \\ 0 & 1 & 0 \\ 0 & 0 & 9\end{bmatrix}$. This is a diagonal covariance matrix.

Extract the standard deviations.

```{r}
Sigma <- diag(c(4, 1, 9))
sqrt(diag(Sigma))                # std devs
```

:::


## Matrix Inverses and Rank

### Inverse

The **inverse** of a square matrix $A$ is a matrix $A^{-1}$ such that $AA^{-1} = A^{-1}A = I$.

- Not all square matrices have an inverse.
- Inverse exists only if matrix is **full rank**.

```{r}
A <- matrix(c(2, 1, 1, 1), 2)
solve(A)                         # A-inverse
A %*% solve(A)                   # should be identity
```

## Full Rank

A matrix has **full rank** if its columns are linearly independent.

- $n \times p$ matrix has full column rank if rank = $p$.
- A square matrix is invertible if and only if full rank.

```{r}
B <- matrix(c(1, 2, 2, 4), 2)
qr(B)$rank # use QR decomposition to find rank
```

## Why Rank Matters

- If $X$ is not full rank, $X'X$ is not invertible.
- In regression, full rank $X$ ensures a unique $\hat\beta$.
- If rank is less than the number of columns, this means on variable is perfectly collinear with another.

## Examples

::: {#exm-matrix3}

Let $A = \begin{bmatrix}4 & 7 \\ 2 & 6\end{bmatrix}$. Compute $A^{-1}$.

```{r}
A <- matrix(c(4, 2, 7, 6), 2)
solve(A)
```

:::

::: {#exm-matrix4}

Let $C = \begin{bmatrix}1 & 2 \\ 2 & 4\end{bmatrix}$. Is $C$ invertible?

```{r}
C <- matrix(c(1, 2, 2, 4), 2)
qr(C)$rank                      # rank < n
```
We can see that the second column is simple 2 times the first column, so the matrix is not full rank and thus not invertible.
:::

## Matrix Calculus (Introductory)

Matrix calculus helps compute gradients and Hessians of functions involving vectors and matrices.

- Used in optimization, MLE, and Bayesian computation
- Especially common in linear models

### Gradients

If $f(x)$ is a **scalar-valued function** of a **vector** $x$, then the **gradient** of $f$ with respect to $x$, denoted $\frac{\partial f}{\partial x}$ or $\nabla_x f$, is a vector containing the partial derivatives of $f$ with respect to each component of $x$, so that 

$$
\frac{\partial f}{\partial x} =
\begin{bmatrix}
\frac{\partial f}{\partial x_1} \\
\frac{\partial f}{\partial x_2} \\
\vdots \\
\frac{\partial f}{\partial x_n}
\end{bmatrix}
$$.

This vector tells us how the function $f(x)$ changes in each coordinate direction, and it points in the direction of steepest ascent.

::: {#exm-grad1}

Let $f(x) = x_1^2 + 3x_2$, where $x = \begin{bmatrix} x_1 \\ x_2 \end{bmatrix}$.

Then the gradient is

$$
\frac{\partial f}{\partial x} =
\begin{bmatrix}
\frac{\partial}{\partial x_1}(x_1^2 + 3x_2) \\
\frac{\partial}{\partial x_2}(x_1^2 + 3x_2)
\end{bmatrix}
=
\begin{bmatrix}
2x_1 \\
3
\end{bmatrix}.
$$

:::


::: {#exm-grad3}

Let $f(x) = x_1^2 + x_1 x_2 + \log(x_3) + e^{x_4}$, where $x = \begin{bmatrix} x_1 \\ x_2 \\ x_3 \\ x_4 \end{bmatrix}$ and $x_3 > 0$.

Then the gradient is

$$
\frac{\partial f}{\partial x} =
\begin{bmatrix}
\frac{\partial}{\partial x_1}(x_1^2 + x_1 x_2 + \log(x_3) + e^{x_4}) \\
\frac{\partial}{\partial x_2}(x_1^2 + x_1 x_2 + \log(x_3) + e^{x_4}) \\
\frac{\partial}{\partial x_3}(x_1^2 + x_1 x_2 + \log(x_3) + e^{x_4}) \\
\frac{\partial}{\partial x_4}(x_1^2 + x_1 x_2 + \log(x_3) + e^{x_4})
\end{bmatrix}
=
\begin{bmatrix}
2x_1 + x_2 \\
x_1 \\
\frac{1}{x_3} \\
e^{x_4}
\end{bmatrix}
$$.

:::



### Hessians

If $f(x)$ is a **scalar-valued function** of a **vector** $x$, then the **Hessian** of $f$ with respect to $x$, denoted $\frac{\partial^2 f}{\partial x \partial x'}$, is an $n \times n$ matrix of second-order partial derivatives:

$$
\frac{\partial^2 f}{\partial x \partial x'} =
\begin{bmatrix}
\frac{\partial^2 f}{\partial x_1^2} & \frac{\partial^2 f}{\partial x_1 \partial x_2} & \cdots & \frac{\partial^2 f}{\partial x_1 \partial x_n} \\
\frac{\partial^2 f}{\partial x_2 \partial x_1} & \frac{\partial^2 f}{\partial x_2^2} & \cdots & \frac{\partial^2 f}{\partial x_2 \partial x_n} \\
\vdots & \vdots & \ddots & \vdots \\
\frac{\partial^2 f}{\partial x_n \partial x_1} & \frac{\partial^2 f}{\partial x_n \partial x_2} & \cdots & \frac{\partial^2 f}{\partial x_n^2}
\end{bmatrix}.
$$

The Hessian describes the curvature of $f(x)$ and is used in second-order optimization methods (like Newton’s method) and in statistical approximations (e.g., Laplace approximation).

::: {#exm-hess1}

Let $f(x) = x_1^2 + 3x_2$, where $x = \begin{bmatrix} x_1 \\ x_2 \end{bmatrix}$.

Then the Hessian is

$$
\frac{\partial^2 f}{\partial x \partial x'} =
\begin{bmatrix}
\frac{\partial^2}{\partial x_1^2}(x_1^2 + 3x_2) & \frac{\partial^2}{\partial x_1 \partial x_2}(x_1^2 + 3x_2) \\
\frac{\partial^2}{\partial x_2 \partial x_1}(x_1^2 + 3x_2) & \frac{\partial^2}{\partial x_2^2}(x_1^2 + 3x_2)
\end{bmatrix}
=
\begin{bmatrix}
2 & 0 \\
0 & 0
\end{bmatrix}.
$$

:::

::: {#exm-hess2}

Let $f(x) = x_1^2 + x_1 x_2 + \log(x_3) + e^{x_4}$, where $x = \begin{bmatrix} x_1 \\ x_2 \\ x_3 \\ x_4 \end{bmatrix}$ and $x_3 > 0$.

Then the Hessian is

$$
\frac{\partial^2 f}{\partial x \partial x'} =
\begin{bmatrix}
2 & 1 & 0 & 0 \\
1 & 0 & 0 & 0 \\
0 & 0 & -\frac{1}{x_3^2} & 0 \\
0 & 0 & 0 & e^{x_4}
\end{bmatrix}.
$$

:::


## Where These Ideas Arise in Modeling

- **Vectors and matrices**: Represent the outcome variable $y$, a matrix of explanatory variables $X$ (usually including a column of ones in the first column), parameters $\beta$, or residuals $\epsilon$.
- **Matrix multiplication**:
  - Linear predictor: $X\beta$
  - OLS and ML estimate of normal-linear model: $\hat{\beta} = (X'X)^{-1}X'y$
- **Transpose**:
  - Quadratic loss: $(y - X\beta)'(y - X\beta)$ 
  - Covariance formulas: $X'\Sigma^{-1}X$
- **Special matrices**:
  - Identity matrix $I$: appears in prior variances, regularization, ridge regression
  - Diagonal matrices: independent variances, prior precision matrices
  - Symmetric matrices: $X'X$, covariance matrices, Hessians
- **Inverses and rank**:
  - Invert $X'X$ to find MLE in linear regression
  - Non-full-rank $X$ causes identifiability issues
- **Matrix calculus**:
  - Score function: gradient of log-likelihood and log-posteriors
  - Hessian matrix: used for Newton-Raphson and Fisher information 

